\documentclass[11pt]{amsart}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{hyperref}
\title{On the rigidity of the minimum-modulus pattern:\\ a circle-rotation reading}
\author{Claude}
\begin{document}
\maketitle

\begin{abstract}
The minimum moduli $\mu_n^{(c)}$ of a family of characteristic polynomials attached to
$\tau(p_n)$ oscillate, and the locations of their extrema obey a strikingly rigid
arithmetic pattern. This note argues that the rigidity is not special to $\tau$: granted a
single dominant complex frequency, it is the universal rigidity of the return times of an
irrational circle rotation (the three-gap theorem), and the specific pattern is a two-term
continued fraction made visible. The genuinely open questions are isolated at the end. The
note is self-contained; the two analytic inputs it borrows are stated as facts in
\S\ref{sec:inputs} and proved in a companion note.
\end{abstract}

\section{Setup and notation}\label{sec:setup}

Let $\tau$ be the Ramanujan tau function, defined by
$\Delta(q)=q\prod_{k\ge1}(1-q^k)^{24}=\sum_{m\ge1}\tau(m)\,q^m$, and let $p_n$ denote the
$n$-th prime. Fix the sequence $h_0=1$, $h_n=\tau(p_n)$ for $n\ge1$.

\emph{The recurrence and its data.} Equation~(D) of the manuscript,
\[
n\,h_n=\sum_{r=1}^{n} j_r\,h_{n-r}\qquad(n\ge1),
\]
is a bijection between sequences $(h_n)_{n\ge0}$ with $h_0=1$ and \emph{data sequences}
$(j_r)_{r\ge1}$. In generating-function form, with $H(t)=\sum_{n\ge0}h_nt^n$ and
$J(t)=\sum_{r\ge1}j_rt^r$, it reads $tH'=HJ$, i.e.\ $J=t\,(\log H)'$.

\emph{The deformation.} For a parameter $c$, prepend $c$ to the data sequence,
$j^{(c)}=(c,j_1,j_2,\dots)$, and run~(D) forward from $h_0=1$ to obtain the deformed
sequence $h^{(c)}=(h_n^{(c)})_{n\ge0}$. Equivalently the generating function is multiplied
by an exponential,
\[
H^{(c)}(t)=e^{ct}\,G(t),\qquad G(t)=\exp\!\int_0^t J(s)\,ds,
\]
so that $h_n^{(c)}=[t^n]e^{ct}G(t)$. Throughout we take $c=1$, the case in the computational
run.

\emph{The spectra.} To each $n$ the manuscript attaches the monic degree-$n$ polynomial
(Theorem~2.3)
\[
\chi_n^{(c)}(x)=\sum_{r=0}^{n}\binom{n}{r}\,r!\,h_r^{(c)}\,(-1)^r\,x^{\,n-r},
\]
and writes $\mu_n^{(c)}=\min\{\,|\zeta|:\chi_n^{(c)}(\zeta)=0\,\}$ for the modulus of its
root nearest the origin. It is the sequence $(\mu_n^{(c)})_n$, plotted against $n$, whose
oscillation is at issue.

\section{The two analytic inputs}\label{sec:inputs}

The argument below borrows exactly two facts. Both are established in the companion
asymptotics note; here they are simply named, with the constants defined as they arise.

\textbf{Input 1 (dominant singularity).} Since $J=tH'/H$, the singularities of $G$ are the
zeros of $H(t)=1+\sum_{n\ge1}\tau(p_n)t^n$. The zero nearest the origin is a
complex-conjugate pair $t_0,\bar t_0$; writing it in polar form
\[
t_0=\rho\,e^{i\theta},\qquad \rho=|t_0|\approx0.0435,\qquad
\theta=\arg t_0\approx0.2432\,\pi,
\]
this defines the modulus $\rho$ and the \emph{argument} $\theta$. (Numerically $t_0$ is
stable under truncation of $H$, and $\rho$ is corroborated by the coefficient growth
$|[t^n]G|^{1/n}\to1/\rho$.)

\textbf{Input 2 (coefficient form and the minima).} By singularity analysis
\cite{FlajoletOdlyzko1990,FlajoletSedgewick2009}, a dominant conjugate pair at
$t_0=\rho e^{i\theta}$ forces the leading-order shape
\[
h_n^{(c)}\;\sim\;C\,\rho^{-n}\,\cos\!\bigl(n\theta+\varphi\bigr),
\]
in which $C>0$ is a real constant and $\varphi$ is a fixed \emph{phase offset} (it absorbs
the argument of the complex amplitude attached to the singularity and the shift contributed
by the $e^{ct}$ factor; only its value, not its origin, matters below). Because
$\chi_n^{(c)}$ is monic, the product of its roots is $n!\,h_n^{(c)}$, and with the bulk of
the spectrum varying smoothly in $n$ the variation of $\mu_n^{(c)}$ tracks $|h_n^{(c)}|$.
Consequently $\mu_n^{(c)}$ dips toward zero exactly at the sign changes of $h_n^{(c)}$,
i.e.\ where the phase
\[
n\theta+\varphi\quad\text{is nearest an odd multiple of}\quad \tfrac{\pi}{2}.
\]

\section{The empirical regularity to be explained}\label{sec:data}

A run to $n=500$, followed by a difference-table analysis of the lower and upper envelopes
of $(\mu_n^{(1)})_n$, reports the following. The successive local minima occur at indices
whose consecutive gaps are drawn from exactly \emph{two} values, $\{4,5\}$; the gap $5$
occurs on a strict period-$9$ schedule; and the second and third differences of the gap
sequence are perfectly periodic with period $9$ (no jitter anywhere). One full block is
\[
8\times 4+1\times 5=37,
\]
containing $9$ minima, so the pattern recurs with period $37$ in $n$; the upper envelope is
the same, phase-shifted. The average gap is $37/9=4.111\ldots$, and an independent Fourier
analysis of the same data places its dominant peak at period $4.10$. So a discrete method
(difference tables) and a spectral method (FFT) agree on a fast period $\approx4.1$ and a
long period $37$.

\section{The strictness is the fingerprint of a circle rotation}

A sinusoid sampled at the integers would not produce gaps from exactly two values with the
$5$ on a strict period-$9$ schedule and perfectly periodic higher differences; it would
jitter. The pattern of \S\ref{sec:data} --- two gap lengths, one inserted quasi-periodically
--- is the signature of a \emph{Sturmian} (mechanical) word
\cite{MorseHedlund1940,Lothaire2002}, and Sturmian words are exactly the symbolic codings of
an irrational rotation of the circle.

Input~2 makes the rotation explicit. The minima of $\mu_n^{(c)}$ fall where the phase
$n\theta+\varphi$ is nearest an odd multiple of $\pi/2$; tracking them is tracking the
returns of the orbit $\{\,n\,(\theta/\pi)+\beta\,\}$ near $\tfrac12$ on the circle, an
equidistributed orbit in the sense of Weyl \cite{Weyl1916,KuipersNiederreiter1974}. The
\emph{three-gap theorem} --- conjectured by Steinhaus, first proved by S\'os, Sur\'anyi, and
\'Swierczkowski \cite{Sos1958,Suranyi1958,Swierczkowski1958,vanRavenstein1988} --- states
that the gaps between successive returns of any irrational rotation take at most three
distinct values, arranged quasi-periodically. Here they take two, $\{4,5\}$. So the rigidity
is not a special fact about $\tau$: it is the universal rigidity of rotation return times.
Once one grants the single dominant frequency $\theta$ of Input~1, the strict two-value
pattern is \emph{forced}; nothing else was available.

\section{The specific pattern is the continued fraction made visible}

The average gap is $37/9=4.111\ldots$, and as a continued fraction
\[
\frac{37}{9}=[4;9]=4+\frac{1}{9}.
\]
That two-term expression \emph{is} the observed law: the integer part $4$ is the base gap;
the partial quotient $9$ is the period of the correction (one extra step --- a $5$ --- every
nine). The minima locations are, to this order, a Beatty sequence
$\lfloor k\,(37/9)+\beta\rfloor$ \cite{Beatty1926,HardyWright}. The fast period $\pi/\theta$
predicted by Input~2 equals $1/0.2432\approx4.11$, matching $37/9$. So the puzzle of why an
$8\times4+5$ block repeats with period $37$ collapses to a two-term continued fraction; the
Fourier peak at $4.10$ is the blurred analytic shadow of this crisp arithmetic object, which
is why the difference-table view looks far more rigid than a spectral peak has any right to.

\section{Where the interesting number theory actually lives}

Two facts remain unexplained by the model.

\emph{First}, that $H(t)=1+\sum_n\tau(p_n)t^n$ has a single, clean, well-isolated dominant
complex zero at all (Input~1). That one conjugate pair should govern everything is a genuine
statement about the analytic structure of a \emph{prime-indexed} $\tau$ series --- an object
with no modularity, no functional equation, and no Euler product. The dominant-singularity
principle \cite{FlajoletOdlyzko1990,FlajoletSedgewick2009} says what such a zero \emph{would}
control, not why this series should have so clean a one. I know no reason.

\emph{Second}, and more pointed: the value of the rotation number $\theta/\pi$ and its
proximity to $9/37$. Is it \emph{exactly} $9/37$ or merely very close? This is decidable, and
the meaning hinges on it.
\begin{itemize}
\item If $\theta/\pi$ is \emph{irrational} with continued fraction $[0;4,9,a_3,\dots]$ and
$a_3$ large, then $9/37$ is an excellent rational approximant, the coding is Sturmian
\cite{MorseHedlund1940,Lothaire2002}, and near $n\sim 37\,a_3$ a \emph{second-order defect}
must appear --- a block with seven or nine $4$'s rather than eight, or two consecutive $5$'s.
The run to $n=500$ shows about thirteen clean cycles with no such defect, which already
forces $a_3$ to be large: $\;|\theta/\pi-9/37|\lesssim 1/(37\cdot500)\approx5\times10^{-5}$.
\item If $\theta/\pi$ were \emph{exactly} $9/37$, the pattern would be exactly period-$37$
forever. That would be astonishing: it would place the nearest zero of a prime-indexed
$\tau$ series at a rational multiple of $\pi$, demanding an arithmetic explanation.
\end{itemize}
My honest bet is the first --- irrational, $[0;4,9,a_3,\dots]$ with $a_3$ large --- because
there is no reason a transcendental-looking $\theta$ assembled from $\tau$ at primes should
be a rational multiple of $\pi$, and every real number masquerades as a small-denominator
rational when known to only four digits. I hold it loosely; the second case, if true, would
be worth a paper.

\section{How to settle it}

Pin $\theta$ down to eight or ten digits (rather than the four currently in hand), form
$\theta/\pi$, and read its continued fraction \cite{HardyWright}. If it begins
$[0;4,9,a_3,\dots]$, then $a_3$ \emph{predicts the exact $n$ at which the first second-order
defect must occur} --- a prediction tested by extending the run past $n=500$ to that point.
A defect at the predicted place proves the rotation-number account and shows $\theta/\pi$
irrational; no defect to all computable precision would be the surprising case. In the
language of circle dynamics this is just reading the rotation number and its convergent
denominators off the orbit \cite{KatokHasselblatt1995}.

\begin{thebibliography}{99}
\bibitem{MorseHedlund1940} M.~Morse and G.~A.~Hedlund, \emph{Symbolic dynamics~II. Sturmian
  trajectories}, Amer. J. Math. \textbf{62} (1940), 1--42.
\bibitem{Lothaire2002} M.~Lothaire, \emph{Algebraic Combinatorics on Words}, Encyclopedia of
  Mathematics and its Applications \textbf{90}, Cambridge University Press, 2002 (Ch.~2,
  Sturmian words).
\bibitem{Sos1958} V.~T.~S\'os, \emph{On the distribution mod~$1$ of the sequence $n\alpha$},
  Ann. Univ. Sci. Budapest. E\"otv\"os Sect. Math. \textbf{1} (1958), 127--134.
\bibitem{Suranyi1958} J.~Sur\'anyi, \emph{\"Uber die Anordnung der Vielfachen einer reellen
  Zahl mod~$1$}, Ann. Univ. Sci. Budapest. E\"otv\"os Sect. Math. \textbf{1} (1958),
  107--111.
\bibitem{Swierczkowski1958} S.~\'Swierczkowski, \emph{On successive settings of an arc on the
  circumference of a circle}, Fund. Math. \textbf{46} (1958), 187--189.
\bibitem{vanRavenstein1988} T.~van Ravenstein, \emph{The three gap theorem (Steinhaus
  conjecture)}, J. Austral. Math. Soc. Ser.~A \textbf{45} (1988), no.~3, 360--370.
\bibitem{Beatty1926} S.~Beatty, \emph{Problem 3173}, Amer. Math. Monthly \textbf{33} (1926),
  159; solutions, \textbf{34} (1927), 159--160.
\bibitem{HardyWright} G.~H.~Hardy and E.~M.~Wright, \emph{An Introduction to the Theory of
  Numbers}, 6th ed., Oxford University Press, 2008 (continued fractions, Ch.~X).
\bibitem{Weyl1916} H.~Weyl, \emph{\"Uber die Gleichverteilung von Zahlen mod.~Eins}, Math.
  Ann. \textbf{77} (1916), 313--352.
\bibitem{KuipersNiederreiter1974} L.~Kuipers and H.~Niederreiter, \emph{Uniform Distribution
  of Sequences}, Wiley, New York, 1974.
\bibitem{KatokHasselblatt1995} A.~Katok and B.~Hasselblatt, \emph{Introduction to the Modern
  Theory of Dynamical Systems}, Cambridge University Press, 1995 (rotation number, circle
  maps).
\bibitem{FlajoletOdlyzko1990} P.~Flajolet and A.~M.~Odlyzko, \emph{Singularity analysis of
  generating functions}, SIAM J. Discrete Math. \textbf{3} (1990), no.~2, 216--240.
\bibitem{FlajoletSedgewick2009} P.~Flajolet and R.~Sedgewick, \emph{Analytic Combinatorics},
  Cambridge University Press, 2009.
\end{thebibliography}

\end{document}
